\textbf{Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated}

### 2022-09-14 submission/manuscript.Rmd
manu

<!-- Context -->
Literature shows that the context surrounding a stimulus often plays an important role in the assessment of its reality [a process henceforth referred to as *simulation monitoring*, @makowski2019phenomenal; @makowski2018cognitive]. The extensive search and processing of cues within ambiguous stimuli being an increasingly complex and cognitively effortful strategy [@michael2021source; @susmann2021persuasion], people tend to draw on peripheral contextual cues (**Figure 1**), such as the source of the stimulus, and its credibility, authority and expertise, to help facilitate their evaluation [petty1986elaboration; @susmann2021persuasion; @michael2021source]. However, the atomization and decontextualization of information allowed by online social media (where text snippets or video excerpts are mass-shared with little context) can render this task difficult. In the absence of contextual information, what drives our beliefs of reality?

@@

<!-- Internal features -->
The evidence shows that inter-individual characteristics play a role in the formation of beliefs of reality, with factors such as cognitive style, prior beliefs, and personality traits significantly impacting simulation monitoring [@bryanov2021determinants; @ecker2022psychological; @sindermann2020short]. For instance, individuals with higher levels of analytical reasoning have been found to better discriminate real from fake stimuli [@pehlivanoglu2021role; @pennycook2019lazy]. Prior knowledge or beliefs about the stimulus influences one's perception of it by biasing the attention deployment towards information in line with our expectations [@britt2019reasoned], and dispositional traits, such as high levels of narcissism and low levels of openness and conscientiousness, have been associated with greater susceptibility to fake news [@piksa2022cognitive; @sindermann2020short].

<!-- Reactive features -->
Beyond stimulus- and individual-related characteristics, evidence suggests that the interaction between the two, i.e., the subjective reaction associated with the experience of a given stimulus, contributes to simulation monitoring decisions. For instance, the intensity of experienced emotions have been shown to increase the sense of presence, i.e., the extent to which one feels like "being there", as if the movie or the VR environment was real [@sanchez2005presence; @makowski2017being], and beliefs that the stimuli were fake led to emotion down-regulation [@makowski2019phenomenal; sperduti2017distinctive]. However, @martel2020reliance reported that heightened stimulus emotionality was associated with greater belief in fake, but not real, information. Other factors, such as the stimulus' perceived self-relevance [@sperduti2016paradox; @goldstein2009pleasure], as well as familiarity  [@begg1992dissociation], could also play a role in our processing and reaction to real vs. non-real material.

<!-- Attractiveness -->
Images of faces, one of the most common artificial intelligence (AI) targets, integrate components of emotional reaction, salience, self-relevance via attractiveness (REF). Talk about possible links between attractiveness and simulation monitoring

<!-- Goals and hypotheses -->

@@

@@

@@

The analysis was carried out using *R 4.2* [**REF**], the *tidyverse* [**REF**], and the *easystats* collection of packages [**REFs for insight, parameters, performance, see**].

>>>>

<!-- Context -->
Literature shows that the context surrounding a stimulus often plays an important role in the assessment of its reality [a process henceforth referred to as *simulation monitoring*, @makowski2019phenomenal; @makowski2018cognitive]. With the extensive search and processing of cues within ambiguous stimuli being an increasingly complex and cognitively effortful strategy [@michael2021source; @susmann2021persuasion], people tend to draw on peripheral contextual cues (**Figure 1**), such as the source of the stimulus, and its credibility, authority and expertise, to help facilitate their evaluation [petty1986elaboration; @susmann2021persuasion; @michael2021source]. However, the atomization and decontextualization of information allowed by online social media (where text snippets or video excerpts are mass-shared with little context) can render this task difficult [*REF*]. In the absence of contextual information, what drives our beliefs of reality?

@@

<!-- Internal features -->
Moreover, evidence from research indicates that inter-individual characteristics also play a crucial role in the formation of beliefs of reality, with factors such as cognitive style, prior beliefs, and personality traits significantly impacting simulation monitoring [@bryanov2021determinants; @ecker2022psychological; @sindermann2020short]. For instance, individuals with higher levels of analytical reasoning have been found to better discriminate real from fake stimuli [@pehlivanoglu2021role; @pennycook2019lazy]. Prior knowledge or beliefs about the stimulus influences one's perception of it by biasing the attention deployment towards information in line with our expectations [@britt2019reasoned], and dispositional traits, such as high levels of narcissism and low levels of openness and conscientiousness, have been associated with greater susceptibility to fake news [@piksa2022cognitive; @sindermann2020short].

<!-- Reactive features -->
Beyond stimulus- and individual-related characteristics, evidence suggests that the interaction between the two, i.e., the subjective reaction associated with the experience of a given stimulus, contributes to simulation monitoring decisions. For instance, the intensity of experienced emotions have been shown to increase one's sense of presence in a simulated reality, i.e., the extent to which one feels like "being there", as if the movie or the VR environment was real [@sanchez2005presence; @makowski2017being], and beliefs that the stimuli were fake were found to result in emotion down-regulation [@makowski2019phenomenal; @sperduti2017distinctive]. In line with these findings, other studies on susceptibility to fake news have also found heightened stimulus emotionality to be associated  with greater belief [@martel2020reliance; @bago2022emotion]. Additionally, other factors, such as the stimulus' perceived self-relevance [@sperduti2016paradox; @goldstein2009pleasure], as well as familiarity [@begg1992dissociation], could also play a role in our processing and reaction to real as opposed to non-real material.

<!-- Attractiveness -->
With the advancement of technology, AI-generated images of faces have increasingly been used to study various cognitive mechanisms and processes, such as that underlying emotion processing [@dyck2008recognition; @dyck2010virtual]. Notably, the ability to manipulate the degree of emotionality and salience of virtual faces, by changing its facial dimensions, make them ideal targets for simulation monitoring research [@fuentes2018evolutionary; @ferstl2018perceptual]. To this effect, the attractiveness of AI-generated faces, which integrates both components of emotional intensity and saliency [@o2003beauty; @debruine2007dissociating; sobieraj2014beautiful], have indeed been associated with its perceived realness. To be specific, research findings suggest that artificially created faces were judged to be less attractive as compared to real faces [@liefooghe2022faces; tsikandilakis2019beauty; @diel2022familiarity]. Interestingly, @liefooghe2022faces further reports that attractiveness ratings differed significantly between participants who were told that the faces were AI-generated from those who had no prior knowledge. Whereas such studies posit that lower perceived realness of a given stimulus leads to lower ratings of attractiveness, little is known about the impacts of attractiveness on simulation monitoring.

<!-- Goals and hypotheses -->
The goal of this study is therefore to determine whether the attractiveness of artificially-generated faces affects our reality judgement of it. Based on the Embodied Reality Theory [@makowski2018cognitive; @makowski2019phenomenal], which suggests that highly salient stimuli are perceived to be more real, we hypothesize a quadratic relationship between the perceived realness and attractiveness of a given virtual stimulus, i.e., faces rated as highly attractive or unattractive will be judged to be more real. Additionally, we further postulate that inter-individual characteristics, such as personality traits and predisposed attitudes, will interact with the relationship between perceived realness and attractiveness. Moreover, given the well-established positive correlation between attractiveness and perceived trustworthiness as well as familiarity often reported in previous studies, we posit that virtual faces judged to be real will also be rated as more trustworthy and familiar [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef; @little2011facial].

In line with open-science standards, all the material (stimuli generation code, experiment code, raw data, analysis script with complementary figures and analyses, preregistration, etc.) is available at [**https://github.com/RealityBending/FakeFace**](https://github.com/RealityBending/FakeFace){.uri}.

@@

109 photos of neutral-expression faces of real individuals were retrieved from the American Multiracial Face Database (AMFD, [@chen2021broadening]). In the first part of the study, participants answered a series of personality questionnaires. This includes the *Mini-IPIP6* [24 items, @sibley2011], to measure 6 "normal" personality traits (Extraversion, Openness, Conscientiousness, Agreeableness, Neuroticism and Honesty-Humility); the *FFNI-BF* [30 items, @jauk2022validation] to measure 9 facets of narcissism (acclaim-seeking, distrust, entitlement, exploitativeness, indifference, lack of empathy, manipulativeness, need for admiration and thrill-seeking); the *R-GPTS* [18 items, @freeman2021revised], to measure 2 sub-scales of paranoid thinking (persecution and reference ideation); the *IUS-12* [12 items, @carleton2007fearing], to assess beliefs about uncertainty and the future; the *SIAS-6* and the *SPS-6* [6 items each, @peters2012development], to assess social anxiety levels as well as a 10-item scale that measures participants general attitudes towards artificial intelligence (of which 5 items were adapted from the *GAAIS* [@schepman2020initial]).

In the second part of this study, the set of facial images retrieved from the AMFD were presented to the participants for 500ms each, in a randomized order. Following each facial display, participants were asked to rate each face on its **Attractiveness** ("I find this person attractive"), **Beauty** ("This face is good-looking"), **Trustworthiness** ("I find this person trustworthy") and **Familiarity** ("This person reminds me of someone I know") using a visual analog scale.

In the last part of the study, participants were informed that half of the facial images previously seen were AI-generated. The same set of facial images displayed before were then presented for a second set of trials in a new randomized order. As per the first set, each facial stimuli was presented for 500ms. After each display, participants were asked to judge whether they thought the face they saw was real (or fake) using a visual analog scale.

This experiment was implemented using *jsPsych* [@de2015jspsych], and the full set of instructions is available in the experiment code.

@@

103 participants were recruited via *Prolific* ([www.prolificacademic.co.uk](www.prolificacademic.co.uk)), a crowd-sourcing platform providing high data quality [@peer2022]. The only inclusion criterion was a fluent proficiency in English to ensure that the experiment instructions would be well-understood. Participants were incentivised with a reward of about \textsterling 7.5 for completing the study, which took about 50 minutes to finish. Demographic variables (age, gender, and ethnicity) were self-reported on a voluntary basis.

Upon inspection of stimuli ratings, attention check responses and total duration spent to complete the personality questionnaires, 3 participants were removed for spending an implausibly short time to finish the questionnaires and having very low correlations between their stimuli ratings and that of the normative distribution.

The final sample included 100 participants (Mean age = 27.9, SD = 8.5, range: [19, 66]; Sex: 48.0% females, 52.0% males, 0.0% other).

@@

The analysis was carried out using *R 4.2* [@RCoreTeam2022], the *tidyverse* [@wickham2019], and the *easystats* collection of packages [@bayestestRArticle; @correlationArticle; @performanceArticle; @insightArticle].
